Extending Height Above Nearest Drainage to Model Multiple Fluvial Sources in Flood Inundation Mapping Applications for the U.S. National Water Model

### 2023-02-11 manuscripts/owp_fim4_2022/sections/appendix.tex
second review responses and changes submitted to coauthors 20230211

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
%
% Step 1: Set the \documentclass
%
%

%% To submit your paper:
\documentclass[draft]{dependencies/agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{dependencies/trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{rotating}

%% added
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{float}

\linenumbers

%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% Add track changes editors
\addeditor{Fernando Aristizabal}

% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\drafttrue

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%
% ie, \journalname{Water Resources Research}

\journalname{Water Resources Research}

% starts document
\begin{document}

%% Title and Authors
%% ------------------------------------------------------------------------ %%
%  Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

% Example: \title{This is a test title}
%
%\title{Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model}
\title{Extending Height Above Nearest Drainage to Model Multiple Fluvial Sources in Flood Inundation Mapping Applications for the U.S. National Water Model}
%
%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
%
%
\authors{Fernando Aristizabal\affil{1,2,3}, Fernando Salas\affil{3}, Gregory Petrochenkov\affil{1,4}, Trevor Grout\affil{1,3}, Brian Avant\affil{1,3}, Bradford Bates\affil{1,3}, Ryan Spies\affil{1,3}, Nick Chadwick\affil{3}, Zachary Wills\affil{3,5}, Jasmeet Judge\affil{2}}
%
%
\affiliation{1}{Lynker, Leesburg, VA, USA}
\affiliation{2}{Center for Remote Sensing, Agricultural and Biological Engineering Department, University of Florida, Gainesville, FL, USA}
\affiliation{3}{National Water Center, Office of Water Prediction, National Weather Service, National Oceanic and Atmospheric Administration, Tuscaloosa, AL, USA}
\affiliation{4}{Hydrologic Applied Innovations Lab, New York Water Science Center, United States Geological Survey, Troy, NY, USA}
\affiliation{5}{Cooperative Institute for Satellite Earth System Studies, University of Maryland, College Park, MD, USA}
%
%(repeat as many times as is necessary)
%
%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)
%
% Example: \correspondingauthor{First and Last Name}{email@address.edu}
%
\correspondingauthor{Fernando Aristizabal}{fernando.aristizabal@noaa.gov}

%% Keypoints
%% Keypoints, final entry on title page.

%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences

% Example:
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
% \end{keypoints}
%
\begin{keypoints}
\item Flood maps derived from Height Above Nearest Drainage (HAND) are subject to nearest drainage line limitations that affect inundation skill.
\item A means of resolving this limitation is provided by reducing HAND processing units to level paths with effective unit stream order.
\item Discretizing the stream network for HAND computation affects the stage-discharge relationship and leads to higher skill inundation.
\end{keypoints}
%

%% Abstract
% ------------------------------------------------------------------------ %%
%
%  ABSTRACT and PLAIN LANGUAGE SUMMARY
%
% A good Abstract will begin with a short description of the problem
% being addressed, briefly describe the new data or analyses, then
% briefly states the main conclusion(s) and how they are supported and
% uncertainties.

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have 
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%

%% \begin{abstract} starts the second page

\begin{abstract}
Height Above Nearest Drainage (HAND), a drainage normalizing terrain index, is a means able of producing flood inundation maps (FIMs) from the National Water Model (NWM) at large scales and high resolutions using reach-averaged synthetic rating curves. 
We highlight here that HAND is limited to producing inundation only when sourced from its nearest drainage line, thus lacks the ability to source inundation from multiple fluvial sources.
A version of HAND, known as Generalized Mainstems (GMS), is proposed that discretizes a target stream network into segments of unit Horton-Strahler stream order known as level paths (LP).
The FIMs associated with each independent LP are then mosaiced together, effectively turning the stream network into discrete groups of homogeneous unit stream order by removing the influence of neighboring tributaries.
Improvement in mapping skill is observed by significantly reducing false negatives at river junctions when the inundation extents are compared to FIMs from that of benchmarks.
A more marginal reduction in the false alarm rate is also observed due to a shift introduced in the stage-discharge relationship by increasing the size of the catchments.
We observe that the improvement of this method applied at 4-5\% of the entire stream network to 100\% of the network is about the same magnitude improvement as going from no drainage order reduction to 4-5\% of the network.
This novel contribution is framed in a new open-source implementation that utilizes the latest combination of hydro-conditioning techniques to enforce drainage and counter limitations in the input data.
\end{abstract}
%
\section*{Plain Language Summary}
Flooding is one of the most impactful natural disasters on life and property.
The United States National Water Model (NWM) provides flood forecasts for the entire country so that adequate warnings can be raised to the public to enable safe evacuations and protective measures.
In order to convert flow rates from the NWM to flood inundation maps (FIM), a model, known as Height Above Nearest Drainage (HAND), is used that converts elevation data from height above mean sea-level to height above the nearest river bottom.
This model suffers from issues in mapping performance because inundation sourced from rivers is only considered from the nearest river line.
We developed a technique that mitigates these errors by removing consideration for neighboring tributaries in the relative elevation computation process.
This is done by splitting the stream network into continuous river segments known as level paths (LPs).
These LPs have no tributaries, thus are known to be stream lines with a unit stream order indicating no branching.
HAND is computed independently for each LP and the resulting FIMs are mosaiced together to form one seamless map.
We compared these HAND derived FIMs to maps from physically-based models and found improvement in mapping performance.
%

%%% Suggested section heads:
% \section{Introduction}
%
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.
%
% \subsection{A descriptive heading about methods}
% More about Methods.
%
% \section{Data} (Or section title might be a descriptive heading about data)
%
% \section{Results} (Or section title might be a descriptive heading about the
% results)
%
% \section{Conclusions}

% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Flooding is one of the most significant natural disasters in the United States (US) affecting both the loss of life and property. 
In 2017 and 2019, river and flash flooding combined represented the leading cause of death and the second leading cause in 2018 among all natural disasters in the US \cite{national_weather_service_2020,national_weather_service_2019,national_weather_service_2018}. 
More than an average of 104 deaths per year are attributed to flood events from the 10 year period ending in 2019 \cite{us_department_of_commerce_2020}. 
With respect to property damages, river and flash flooding have contributed to 60.7, 1.6, and 3.7 billion non-inflation adjusted US dollars in the annual periods of 2017 to 2019, respectively \cite{national_weather_service_2020,national_weather_service_2019,national_weather_service_2018}, with the large spike in 2017 attributed to the Hurricane Harvey event along the Gulf Coast. 
Trends related to flood damages and fatalities have been steadily increasing over recent decades \cite{mallakpour2015changing,downton2005reanalysis,kunkel1999temporal,pielke2000precipitation,corringham2019effect}. 
Some are expecting that the hydrologic cycle will intensify due to climate change which will lead to more extreme precipitation in some areas along with a greater risk of flooding \cite{tabari2020climate,milly2002increasing,wing2018estimates}. 
Increasing trends in frequency and risk are not uniform across spatial regions with work by \citeA{slater2016recent} indicating that trends are increasing across the US Midwest and Great Lakes regions while decreasing in the coastal Southeast, Southwest, and California.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Forecasting}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Operational flood forecasting systems are primary tools in developing accurate forecasts for public awareness prior to life threatening and property damaging events. 
One of these operational systems is the Advanced Hydrologic Prediction System (AHPS) maintained by the National Oceanic Atmospheric Administration (NOAA) National Weather Service (NWS) with thousands of forecasting points across the US at typically short forecast horizons of 24 or 72 hours \cite{mcenery2005noaa}.
AHPS utilizes a variety of techniques including distributed modeling \cite{reed2004overall,koren2004hydrology,duan2003global}, ensemble forecasting \cite{day1985extended,seo2000simulation,mullusky2002simplified,herr2002simplified}, hydraulic routing \cite{cajina2002fldview,fread1973technique}, multi-sensor precipitation methods \cite{seo2002real,kondragunta2001outlier,breidenbach1999accounting,bonnin1996noaa}, flood inundation mapping \cite{cajina2002fldview}, and calibration techniques \cite{zhang2003hydrologic,hogue2003multi,duan2003global,gupta2003advances,parada2003multi}.
AHPS provides forecasting services in the form of ensemble streamflows at more than 3,600 locations and flood inundation maps (FIM) at more than 150 of those points shown in Figure \ref{fig:forecast_points}.
Additionally, two forecasting networks, Full Resolution (FR) and Mainstems (MS) stream networks, relevant to the National Water Model (NWM) (see Section \ref{ssec:national_water_model}) are rendered in Figure \ref{fig:forecast_points}.
The FR network refers to the entire NWM forecasting domain while MS refers to the subset of the FR network that is at or downstream of AHPS forecasting points (see Section \ref{ssec:national_water_model}).
On an approximate basis, there is only one forecast point every 1,450 km of river (FR) and one forecast point with FIM every 29,000 km of river (FR).
Despite the AHPS advances in operational flood forecasting, it lacks sufficient domain coverage, spatial resolution, and long-range forecast horizons to address the increasingly complex water challenges facing the US.
%
\begin{figure}[H]
\centering
\includegraphics[scale=0.75]{figures/forecast_points.jpg}
\caption{Forecast points with and without Flood Inundation Maps (FIM) in United States' Advanced Hydrologic Prediction System (AHPS).
Note that only a small fraction of the AHPS forecast points have existing FIM.
Also shown are the National Water Model (NWM) stream networks at the Full Resolution (FR) and Mainstems (MS) resolution.
The FR network constitutes the entire NWM stream network while the MS resolution network is the FR network at or downstream of the AHPS forecast points shown.
}
\label{fig:forecast_points}
\end{figure}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{National Water Model}
\label{ssec:national_water_model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Additional work is required to address the gaps that the AHPS leaves in terms of spatial resolution, spatial coverage, and temporal forecast horizons.
In response to growing stakeholder demand for enhanced and integrated water resource forecasts, the Office of Water Prediction (OWP) at the National Water Center (NWC) along with its partners at the National Center for Atmospheric Research (NCAR) have developed and implemented operationally the NWM which is a configuration of the Weather Research and Forecasting Hydrologic Model (WRF-Hydro) \cite{salas2018towards,gochis2021wrf,cosgrove2019evolution}. 
The NWM forecasts river discharges at more than 2.7 million forecast points at a variety of time horizons including lookback-range (3-28 hrs), short-range (18 hr), medium-range (10 day) and long-range (30 day) forecast horizons.
The NWM enhances the spatial and temporal domain of the current AHPS capabilities operated at the 13 River Forecast Centers (RFC) in areas known as `hydro-blind'.
As a complement to the operational NWM, RFC forecasts from AHPS forecast points are assimilated in the NWM and routed downstream to the next downstream AHPS forecast point where the process iterates again.
This assimilation into the NWM is used to enhance forecasting skill by leveraging best available regional-scale forecasts.
The river network upon which this special assimilation technique operates on is herein referred to as the Mainstem (MS) stream network.
Figure \ref{fig:forecast_points} shows the NWM V2.1 FR stream network as well as the NWM V2.1 MS network.
The MS network contains roughly 120 thousand forecasting points or roughly 4.4\% of the reaches of the FR stream network.

The National Hydrography Dataset Plus (NHDPlus) V2.1 is the basis for the ``hydrofabric'' in the NWM due to its comprehensive use with the hydrologic communities' stakeholders \cite{mckay2012nhdplus,nhdplus2022vectors}. 
The term ``hydrofabric'' is used within the NWM jargon to describe the subset of hydrography composed of the geospatial datasets required for hydrologic modeling including but not limited to stream networks, catchments, channel properties, and elevation data. 
The NWM provides stream forecasts at these hydrofabric segments using the Muskingum-Cunge method to reduce computational requirements of a continental scale model but fails to consider backwater dynamics \cite{bedient2008hydrology,ponce1994variable,gochis2021wrf}.
The need for high resolution FIM at 10 m or better requires additional post-processing from the principal output of the NWM which is forecast river discharges at the reach scale.
The use of a 2-dimensional (2D) hydrodynamic model across a continental-scale and high spatial resolutions is very cost prohibitive especially in an operational setting.
The Height Above Nearest Drainage (HAND) terrain model is one such technique that can be used, along with synthetic rating curves (SRC), to convert 1-dimensional (1D) riverine discharges to stages, and finally to inundation extents.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Height Above Nearest Drainage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
HAND normalizes topography along the nearest drainage path and it has been demonstrated to be a good proxy and indicator of a series of important environmental conditions including soil environments, landscape classes, soil gravitational potentials, geomorphologies, soil moisture, and groundwater dynamics \cite{renno2008hand,nobre2011height}. 
\citeA{nobre2016hand} showed evidence for utilizing the drainage normalizing HAND dataset as a proxy for flood potential to make static flood inundation maps from known stages.
The terrain index also provides additional utility in the observation of riverine flood inundation mapping from remote sensing especially in areas of high electromagnetic interference such as vegetated and anthropogenic areas \cite{aristizabal2020high,shastry2019using,huang2017comparison,twele2016sentinel,aristizabal2021mapping}.
\citeA{zheng2018river} developed a methodology for determining stage-discharge relationships known as SRCs by sampling reach-averaged parameters from HAND datasets and inputting into the Manning's equation \cite{gauckler1867etudes,manning1890flow}.
This collection of methods, coupling HAND with SRCs, have been experimented with and compared to other sources of FIM including engineering scale models, in-situ observation, and remote sensing based observation with solid results in large spatial scale applications \cite{godbout2019error,johnson2019integrated,garousi2019terrain,nobre2016hand,afshari2018comparison,zheng2018geoflood,teng2015rapid,teng2017flood,zhang2018comparative,hocini2020performance}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HAND's Assumptions and Limitations}
\label{sssec:hand_assumptions_limitations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
HAND operates on many underlying assumptions since it can only be used as an inundation proxy or no physics model and thus, not a true hydrodynamic inundation model \cite{nobre2016hand,liu2016cybergis,liu2020height}.
HAND, to our knowledge, has only been applied to natural, inland, and riverine inundation applications thus it is also missing pluvial, coastal, ground water, and dam break components among other possible sources of flooding.
Additionally, in order to flood an area, HAND assumes all areas eligible for inundation must drain to some nearest stream line which is used for catchment allocation and relative elevation calculation \cite{nobre2016hand,nobre2011height,liu2016cybergis,liu2020height,maidment2017conceptual,garousi2019terrain,zheng2018river,zheng2018geoflood,johnson2019integrated,renno2008hand,hocini2020performance}. 
Stream thalweg networks must also collectively drain to a singular outlet point for a given processing region \cite{nobre2016hand,zheng2018geoflood,renno2008hand}.
Since elevations don't naturally do this, they must undergo a long series of hydro-conditioning processes to enforce monotonically decreasing elevations across an entire processing unit along with hydrologically correct directions of flow \cite{nobre2016hand,nobre2011height,liu2016cybergis,liu2020height,donchyts2016global,renno2008hand}.
The level of digital elevation map (DEM) manipulation required to enforce this assumption can be substantial depending on the region and can be a significant source of error.
The drainage enforcing assumption also interacts with an inability to properly account for fluvial inundation in regions of DEM depressions that lack natural drainage to riverine areas \cite{nobre2016hand,renno2008hand}.

When used for FIM applications, HAND assumes only fluvial inundation sourced from its nearest drainage line is accounted for \cite{nobre2016hand,mcgehee2016modified}.
Catchments are independent of one another for FIM purposes meaning a reaches' stage value is only used to threshold the HAND values within its respective catchment \cite{liu2016cybergis,zheng2018river,zheng2018geoflood}.
This assumption plays to the ``Nearest Drainage'' term in HAND and creates a significant limitation within HAND for FIM applications \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
At the junction of high stream order and high flow rivers with lower flow tributaries, there can be a lack of inundation extents exhibited which is known colloquially in the forecasting community as the ``catchment boundary problem''.
The academic community has somewhat referenced this issue before but it has been characterized more as a problem with the stream delineation process that comes from thresholding the drainage accumulation maps \cite{nobre2016hand,li2020evaluation}.
Later in this study, we will re-introduce this problem and demonstrate how we initialize with a stream network (that of the NWM's) and thus avoid having to threshold accumulations to some arbitrary value to define stream networks.
We illustrate how computing HAND independently for stream lines of unit stream order can significantly enhance FIM performance by accounting for multiple sources of fluvial inundation that may exist in certain regions and flow scenarios.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HAND Implementations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Due to significant advances in high performance computing (HPC) and large scale high resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (approximately 10 m depending on latitude) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a HPC \cite{maidment2017conceptual,liu2016cybergis}. 
NFIE used open-source dependencies including the Terrain Analysis Using Digital Elevation Models (TauDEM) \cite{tarboton2005terrain} and the Geospatial Data Abstraction Library (GDAL) \cite{warmerdam2008geospatial} to compute HAND for the Continental United States (CONUS) at 331 Hydrologic Unit Code (HUC) 6 processing units in 1.34 central processing unit (CPU) years.
By allocating 31 nodes at 20 cores per node for a total of 620 available cores to the overall operation, it enabled the production to finish up in 36 hours consuming 3.2 terrabyte (TB) of peak memory and 5 TB of total disk space.
Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (NHDPlusHR) which better agrees with the 10 m DEM from the NHDPlusHR program \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstration to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.
Further work by \citeA{djokic2019arc}, implemented a series of improvements to HAND including equidistant reaches, updates to use with NHDPlusHR hydrography, and AGREE DEM reconditioning \cite{hellweger1997agree} into an ESRI Arc-Hydro workflow with use in ArcGIS. 
More notably the software added the ability to derive HAND on both the NWM FR and MS stream networks to consider multiple sources of fluvial inundation along high impact rivers of primary forecasting concern.

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
\label{ssec:owp_fim}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages and inundation extents.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.

Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
HAND produced FIMs are limited in only providing inundation sourced from the nearest drainage line, however, depending on flow conditions and topography, a given area may have multiple contributing fluvial sources of surface water inundation.
The forecasting community, in reviewing HAND, have noted significant negative effects at the confluence of lower flow tributaries with higher flow rivers for which the phrase ``catchment boundary issue'' has been termed.
In previous studies, FIM skill has been shown to be sensitive to the drainage density of the stream network employed as the datum for HAND which is closely related to the maximum Horton-Strahler stream order of the network \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is partly in due to the limitation that catchment boundaries place on inundation extents where only the nearest drainage line can source inundation for any particular area.

Figure \ref{fig:catchment_boundaries_issue} illustrates the exact situation our solution proposes to address where two tributaries converge with a higher order stream segment. 
An actual map with OWP FIM is generated using the NWM full-resolution stream network and compared with a FEMA 100 yr extent (see Section \ref{ssec:evaluation} for more details) showing significant under-prediction in inundation extent.
The higher discharge along the main segment in Figure \ref{fig:catchment_boundaries_issue} of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 
This is due to a lack of representation of backwater conditions in the hydraulic routing techniques used.
As a parallel problem, there is excess water accumulated along the mainstem that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the mainstem catchments.
%
\begin{figure}[H]
\centering
\includegraphics[scale=1.0]{figures/catchment_boundaries_issue.jpg}
\caption{The figure represents an agreement map between a HAND derived FIM and one produced from the Base Level Engineering (BLE) program for a 100 yr magnitude event at HUC8 12090301.
Agreement maps are symbolized by false negatives (FN), true negatives (TN), false positives (FP), and true positives (TP) where inundated represents the positive condition (see Section \ref{ssec:evaluation} for more details).
The streamflows associated with each river segment are shown in CMS while the flow directions are symbolized as red arrows.
The presence of FNs at the confluence of tributaries (circled in red) with the main segment is associated with lower flow rates in the tributaries that don't account for backwater effects.
Additionally, the flow of 1900 CMS from the main segment cannot extend to the neighboring catchments belonging to its tributaries shown here.
Water pools up vertically along the catchment boundaries of the higher order segment distorting rating curve behavior (Section \ref{sec:discussion}).
Sourcing fluvial inundation from HAND is limited to only its nearest drainage line which is the main issue this study aims to address.
}
\label{fig:catchment_boundaries_issue}
\end{figure}
%

We seek to resolve this catchment boundary problem or nearest drainage limitation by discretizing the target stream network into stream networks of reduced, unit stream order to avoid the constraining of catchments by those belonging to lower order neighbors.
By discretizing the network into stream networks of unit stream order (later defined as level paths or LPs), we remove the influence of neighboring catchments that constrain the inundation extent.
This creates much larger and overlapping catchments that can source fluvial inundation from multiple reaches as required by the given river stage at current flow conditions.
We present two successive methods, National Weather Service MS (Section \ref{sssec:nws_mainstems}) and Generalized Mainstems (GMS) (Section \ref{sssec:generalized_mainstems}), implemented that reduce the effective Horton-Strahler stream orders of the networks employed and test our presented hypothesis that stream networks of unit stream order enhance quality of FIM extents produced with HAND by expanding the nearest drainage definition to increase potential inundated areas.

Here we demonstrate how reducing a HAND processing unit's stream network into discrete LPs of singular, effective stream order, can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree GMS (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM.
%
 %
% Materials and Methods
\clearpage % this clears figures before references
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
\label{sec:materials_and_methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM is a fully operational pipeline of software tools to help acquire datasets, cache hydrofabrics, produce FIMs, and evaluate results.
Figure \ref{fig:methods_overview} gives a high level overview of the methodology used in OWP FIM and in this study.
Input data from multiple sources are preprocessed (not illustrated) and then subset to processing areas based on the model used, FR, MS, or GMS.
The standard processing unit of OWP FIM is a HUC8 and the entire NWM FR stream network is used for enforcment.
Later, we explain how only the NWM MS stream network is used for the MS version of HAND, while for GMS, the FR stream network is discretized into LPs before computing HAND.
A series of hydro-conditioning steps enforces the location of stream lines, monotonically decreasing elevations, excavated bathymetry, stream thalweg breaching, and levee enforcement.
After a DEM suitable for HAND's assumptions is conditioned, the FIM hydrofabric is generated including stream network, catchments, HAND, SRCs, and cross-walk table.
The FIM hydrofabric is defined as the datasets required to make an inundation map from discharges including the relative elevation model (REM) or HAND grid, the catchments in vector and raster form, and the hydro-table (contains SRC and cross-walk information).
In operational circumstances, the NWM streamflows are used in conjunction with the FIM hydrofabric to derive forecast FIMs.
However for evaluation purposes, we use the streamflows from the cross-sections of our benchmark model.
As later discussed in Section \ref{ssec:inundation_mapping}, independent FIMs from multiple fluvial sources are mosaiced together.
For the case of MS, two sources are mosaiced together (FR and MS) while for GMS the inundation from every LP is composited together.
The evaluation FIM extents are compared to the extents of the benchmark model and metrics are computed.
%
\begin{figure}[H]
\centering
\includegraphics[scale=6.0]{figures/methods_overview.jpg}
\caption{
Methodology overview detailing high level steps followed in the study.
The flow chart begins with the input data organized by source.
Subsetting the data into processing units depends on which model is being considered.
FR utilizes the entire NWM stream network processed at HUC8 processing areas.
MS only computes HAND using the NWM stream at or downstream of legacy forecasting points.
The resulting inundation from the MS HAND is eventually layered with the FIM from FR HAND to account for high levels of inundation contributed by the mainstem.
Generalized Mainstem (GMS) discretized NWM streams into level paths (LP) then computes HAND and the FIMs independently only to mosaic them later. 
This better accounts for multiple possible sources of fluvial inundation.
The dotted lines denote the use of NWM streamflow forecasts to produce operational FIM but not used in this study.
All acronyms used in the figure are defined in the paper.
}
\label{fig:methods_overview}
\end{figure}
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Dependencies and Architecture}
\label{ssec:software}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM exclusively utilizes free and open source software dependencies including Python 3, GDAL, TauDEM, Geographic Resource Analysis Support System (GRASS), GNU Parallel, and MPICH \cite{python382,gdal2020,tarboton2005terrain,grass2020,tange2015gnu,amer2021mpich}.
Within the Python 3 ecosystem, many common packages are employed including but not limited to RichDEM, GeoPandas, Rasterio, Rasterstats, and Numba \cite{barnes2018richdem,jordahl2014geopandas,lam2015numba}. 
To simplify setup and enhance portability across host operating systems, OWP FIM packages all dependencies up in a Docker image \cite{merkel2014docker}. 
A user only needs to install Docker on their host machine and build the image from the provided recipe. 
Source code is made available for this project on GitHub where a user could consult the Readme.md page for more information on how to acquire the datasets and reproduce the pipeline \cite{inundationMapping2022}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\label{ssec:datasets}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Data sources used within OWP FIM are publicly available from a variety of government sources including the USGS, NWC, Federal Emergency Management Agency (FEMA), and US Army Core of Engineers (USACE) to enhance reproducibility and collaboration among government, academia, and industry.
Instructions for accessing data processed for OWP FIM are provided on the project's GitHub page via an Amazon Web Services (AWS) S3 bucket furnished by the Earth Science Information Partners (ESIP) \cite{esipData2022}.
The National Hydrography Dataset Plus High Resolution (NHDPlusHR) Beta Version is the latest hydrography dataset used for land surface hydrologic modeling in the US \cite{moore2019user}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Stream Network Enforcement}
\label{sec:app_stream_network_enforcement}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The location of the stream network is enforced to ensure general agreement with the NWM network which is used for forecasting the streamflow inputs.
The NHDPlusHR Beta BurnLineEvent layer is used to enforce stream locations in the NHDPlusHR workflow and best agrees with thalweg locations in the DEM used so it is also used here for hydro-enforcement \cite{moore2019user}.
